### 2021-07-07 rad-icfp-2021/correctness.tex
writing section 5

It is standard to give a denotational semantics of a first-order language in a Cartesian category.
An alternative approach consists of giving semantics in a multicategory.
A multicategory generalizes a category by allowing multimorphisms, that is, morphisms from a list of objects to an object.
Most categorical structures from category theory can be phrases similarly in multicategories.

In terms of semantics, a term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
Substitution is still interpreted as composition.

One can consider a syntactic model for a language, which consists in a free multicategory.
Our source and target languages induce syntactic multicategories as follows.

@@

As is standard, $\SynSource$ satisfies the following universal property: 
for every Cartesian multicategory $\catC$ with arrays map and reduce,
and every object $F(\reals)\in\catC$ and morphisms $F(\underline{c})\in\catC(1;F(\reals))$, 
$F(op1)\in\catC(F(\reals);F(\reals))$, $F(op2)\in\catC(F(\reals),F(\reals);F(\reals))$, there is a unique
multifunctor $F:\SynSource\to\catC$ respecting the interpretation and preserving all the structure.

We can, for instance, use this to give a simple semantics of Source 
in the multicategory of Cartesian spaces and smooth maps between them.

\begin{definition}[$\CartSp$]
Let $\CartSp$ be the Cartesian category whose objects are Euclidean spaces
and whose morphisms are smooth functions between Euclidean spaces.
TODO:it's not a multicategory yet

We can interpret Source in $\CartSp$ as follows.
    \begin{tabular}{r c l}
    $\seml \reals \semr$ & $\defeq$& $\RR$ \\
    $\seml$!T1xT2!$\semr$ & $\defeq$& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ 
    \end{tabular}
A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\prod_{1\leq i \leq n}\sem{A_i}$.
Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$. This is
routinely defined by induction on the structure of typing derivations. Constants
!c! $\reals$ are interpreted as constant functions; and the primitives operations !op1,op2!
are interpreted by composing with the corresponding functions, which are smooth. 
Variables are interpreted as $\sem{x_i}(\rho)\defeq \rho_i$. 
The remaining constructions are interpreted as follows.
\begin{tabular}{r c l}
    $\seml$!map2 (x,y.e1) e2 e3!$\semr$ &=& TODO\\
    $\seml$!reduce (x,y.e1) e2 e3!$\semr$ &=& TODO
\end{tabular}
\end{definition}

TODO: what about semantics Target?

@@

One main reason for introducing UNF is to have a better handle over
the computation flow of the term, similar to what ANF or CPS provide. 
A convenient categorical setting for this is to use string diagrams.
To better fit the standard denotational semantics of languages, 
we can instead use concategories.

Concategories are a generalization of categories and multicategories in that 
they allow morphisms both from and to a list of objects. 
Similarly to categories and multicategories, 
most categorical abstractions used for the semantics of languages are easily adapted to them.

In particular, one can form a syntactic concategory, 
which satisfies a similar universal property as syntactic multicategories.
Two such examples are of particular interest for us, 
because they will allow us to interpret Source and Target UNF, 
but also explain the UNF and $UNF^{-1}$ transformations.

\begin{definition}[$\ConcatS$]
Let $\ConcatS$ the syntactic Cartesian concategory whose types are those of Source and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$
and $reduce_{x,y.e1;e2}$.
TODO: actually I just need projection and a representing object for pairs, it might be weaker than Cartesian? edit: I think I need Cartesian.
\end{definition}

@@

\begin{definition}[$\ConcatT$]
Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
with primitives given by 
$op1: \reals \to \reals$, 
$op2: \reals,\reals \to \reals$, 
$map2_{x,y.e}: \Array{\reals}{n},\Array{\reals}{n}\to \Array{\reals}{n}$, 
$reduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
$J^Top1: \reals \to \reals$, 
$J^Top2: \reals \to \reals, \reals$, 
$J^Tmap2_{x,y.e}: \Array{\reals}{n} \to \Array{\reals}{n}, \Array{\reals}{n}$, 
$J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
TODO: not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?
\end{definition}

We can interpret Target UNF in $\ConcatT$ as follows. 
The part common with Source UNF is interpreted in the same way as for Source UNF.

@@

@@

First, we start from the correctness of $\Dsynrevsymbol_{\rho}$:Source UNF $\to$ Target UNF, which is easy to establish, 
and then propagate this information to Source and Target via the $\UNFSymbol, \UNFSymbol^{-1}$ transformations.

\begin{proposition}[Correctness $\Dsynrevsymbol_{\rho}$]
    For every term !$\reals^{\times n} \vdash$ e: $\reals^{\times n+1}$! in Source UNF,\\ 
    $\pi_2 \seml \Dsynrevsymbol_{\rho}$!e!$\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}\seml$!e!$\semr$.
\end{proposition}

>>>>

A multicategory generalizes a category by allowing multimorphisms, that is, morphisms from a list of objects to an object.
Most categorical structures from category theory can be phrased similarly in multicategories.
It is standard to give a denotational semantics of a first-order language in a Cartesian category, and alternatively in a multicategory.

A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
free multicategy on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.

@@

$\SynSource$ satisfies the following universal property: 
for every Cartesian multicategory $\catC$,
and every object $F(\reals)\in\catC$, morphisms $F(\underline{c})\in\catC(1;F(\reals))$, 
$F(op1)\in\catC(F(\reals);F(\reals))$, $F(op2)\in\catC(F(\reals),F(\reals);F(\reals))$, 
there is a unique multifunctor $F:\SynSource\to\catC$ respecting the interpretation and preserving all the categorical structure.

This allows us to give a simple semantics of Source 
in the multicategory of Cartesian spaces and smooth maps between them.

\begin{definition}[$\CartSp$]
Let $\CartSp$ be the Cartesian multicategory whose objects are Euclidean spaces
and whose morphisms $[A_1,\ldots,A_n]\to B$ are smooth functions $A_1\times\ldots A_n \to B$.

We interpret Source in $\CartSp$ as follows. A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\prod_{1\leq i \leq n}\sem{A_i}$.
Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$.

    \begin{tabular}{r c l}
    $\seml \reals \semr$ & $\defeq$& $\RR$ \\
    $\seml$!T1xT2!$\semr$ & $\defeq$& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ 
    \end{tabular}
    \begin{tabular}{r c l}
        !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ \\
        !$\seml$op2$\semr$! & $\defeq$ & op2: $\reals\times \reals\to\reals$ \\
        !$\seml$c$\semr$! & $\defeq$ & $c\in\RR$ \\
        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & \\
        !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ &
    \end{tabular}

Variables are interpreted as projections, let binding uses composition in the multicategory. 
Pairs are interpreted using the Cartesian structure of the multicategory.
\end{definition}

We can similarly interpret Target in a multicategory of smooth-like spaces and functions. 
However, Target is higher-order but $\CartSp$ is not Cartesian Closed. 
Instead, we can for instance interpret Target in the category of Diffeological spaces, 
as in \cite{huot2020correctness}. 
Diffeological spaces (\cite{iglesias2013diffeology}) are a conservative extension of $\CartSp$. 
The key idea will be that a higher order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
to build first order functions that are not smooth.

\begin{definition}
	A \emph{diffeological space} $(X,\plots{X})$ consists of a set $X$ together with, for each $n$ and each open subset $U$ of $\RR^n$,  a set $\plots{X}^U\subseteq [U\to X]$ of functions, called \emph{plots}, such that
	\begin{itemize}
	 	\item all constant functions are plots;
	 	\item if $f:V\to U$ is a smooth function and $p\in\plots{X}^U$, then $f;p\in\plots{X}^V$;
     \item if $\seq[i\in I]{p_i\in\plots{X}^{U_i}}$ is a compatible family of plots $(x\in U_i\cap U_j\Rightarrow p_i(x)=p_j(x))$
     and $\seq[i\in I]{U_i}$ covers $U$,
     then the gluing $p:U\to X:x\in U_i\mapsto p_i(x)$ is a plot.
	 \end{itemize} 
\end{definition}
We call a function $f:X\to Y$ between diffeological spaces \emph{smooth} if, for all plots
$p\in\plots{X}^U$, we have that $p;f\in \plots{Y}^U$. We write $\Diff(X,Y)$ for the set of smooth maps from $X$ to $Y$. 
Smooth functions compose, and so we have a category $\Diff$ of diffeological spaces and smooth functions.

A diffeological space is thus a set equipped with structure.
Many constructions of sets carry over straightforwardly to diffeological spaces.

\begin{example}[Cartesian diffeologies]\label{ex:cartesian-diffeologies}
Each open subset $U$ of $\RR^n$ can be given the structure of a diffeological space by taking all the
smooth functions $V\to U$ as $\plots{U}^V$.
It is easily seen that smooth functions from $V\to U$ in the traditional sense coincide with
smooth functions in the sense of diffeological spaces.
Thus diffeological spaces have a profound relationship with ordinary calculus.
\end{example}

\begin{example}[Product diffeologies]
Given a family $\seq[i\in I]{X_i}$ of diffeological spaces,
we can equip the product $\prod_{i\in I}X_i$ of sets with the
\emph{product diffeology} in which $U$-plots are precisely the functions
of the form $\seq[i\in I]{p_i}$ for $p_i\in\plots{X_i}^U$.  
\end{example}

\begin{example}[Functional diffeology]
We can equip the set $\Diff(X,Y)$ of smooth functions between diffeological spaces with the \emph{functional diffeology}
in which $U$-plots consist of functions $f:U\to \Diff(X,Y)$ such that 
$(u,x)\mapsto f(u)(x)$ is an element of $\Diff(U\times X, Y)$.
\end{example}

We can thus interpret function types !$\seml$A -> B$\semr$! = $\Diff$($\seml$!A!$\semr$,$\seml$!B!$\semr$).

@@

One main reason for introducing UNF is to have a better handle over
the computation flow of the term, similarly to what the ANF or CPS transformations do.
A convenient categorical setting for this is to use string diagrams.
To better fit the standard denotational semantics of languages, 
we can instead use concategories.

Concategories are a generalization of categories and multicategories in that 
they allow morphisms both from and to a list of objects. 
Similarly to categories and multicategories, 
most categorical construtions used for the semantics of functional languages 
have equivalent in concategories.

In particular, one can form a syntactic concategory on some base types and primitives, 
which satisfies a similar universal property as syntactic multicategories.
Two partiular examples are of interest for us, as they will allow us to interpret Source and Target UNF
but they will also help us explain the UNF and $UNF^{-1}$ transformations.

\begin{definition}[$\ConcatS$]
Let $\ConcatS$ be the syntactic Cartesian concategory whose types are those of Source and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$, 
TODO: $map2_{x,y.e}$and $reduce_{x,y.e1;e2}$.
\end{definition}

@@

\begin{definition}[$\ConcatT$]
Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
whose primitives are given by 
$op1: \reals \to \reals$, 
$op2: \reals,\reals \to \reals$, 
$map2_{x,y.e}: \Array{\reals}{n},\Array{\reals}{n}\to \Array{\reals}{n}$, 
$reduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
$J^Top1: \reals \to \reals$, 
$J^Top2: \reals \to \reals, \reals$, 
$J^Tmap2_{x,y.e}: \Array{\reals}{n} \to \Array{\reals}{n}, \Array{\reals}{n}$, 
$J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$ 
\end{definition}

We can interpret Target UNF in $\ConcatT$ as follows. 
The part common with Source UNF is interpreted in the same way as for the case of Source UNF.

@@

In summary, we have the following picture

\[
\begin{tikzcd}
    SourceUNF \ar[r,"\sem{-}"] & Concat1 \ar[r,"\Dsynrevsymbol_{\rho}"] & Concat2 \ar[r,"\UNFSymbol^{-1}"] & C_{Target} \\
    Source \ar[r,"\UNFSymbol"] & C_{Source} \ar[u, hook] & TargetUNF \ar[u,"\sem{-}"] & Target \ar[u, hook] 
\end{tikzcd}
\]

@@

First, we start from the correctness of $\Dsynrevsymbol_{\rho}$:Source UNF $\to$ Target UNF, which is easy to establish, 
and then propagate this information to Source and Target via the $\UNFSymbol$ and $\UNFSymbol^{-1}$ transformations.

\begin{proposition}[Correctness $\Dsynrevsymbol_{\rho}$]
    For every term !$\reals^{\times n} \vdash$ e: $\reals^{\times n+1}$! in Source UNF,
    \begin{center}
        $\pi_2 \seml \Dsynrevsymbol_{\rho}$!e!$\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}\seml$!e!$\semr$
    \end{center} 
\end{proposition}
